Process-based phenotypes: an application to the ontology of cell physiology

### 2011-11-26 trunk/paper/drinking.tex

\section{Introduction}

problem of phenotypes...

why reasoning, why formal, success,...

solved: structures, qualities

open: process phenotypes very complex...  
physiology...
measurements...

@@

@@

CPO contains both structural abnormalities and physiological
abnormalities.

Phenotypes in the CPO are either based on structural abnormalities or
abnormal physiology involving cells or cell components. Structural
abnormalities in the CPO are based on GO's Cellular Component (GO-CC)
hierarchy. GO-CC contains 2,918 classes for cell parts (including {\em
  Cell}) and extracellular components of cells. For each cellular
component class $C$ in the GO-CC, we create a new class labelled {\em
  $C$ phenotype} in the CPO. For example, for the class {\em
  Mitochondrion} ({\tt GO:0005739}) in the GO-CC, we create the class
{\em Mitochondrion phenotype}.

>>>>

\section{Introduction}
Phenotype studies on all scales and levels of granularity are now an
invaluable tool for functional genomics research. Phenotypes of
targeted mutations in animal models are now systematically recorded to
reveal the role of individual genes within a biological system. These
phenotype studies now play a key role in translational research and
are being used to reveal candidate genes for orphan diseases and to
identify chemicals that may have effects on these diseases \cite{}.

The large volume and diversity of phenotypes within different species
and across multiple scales and levels of granularity necessitates the
application of flexible strategies for managing and integrating data
so that it becomes amenable to automated comparative analyses. To
integrate biomedical data across heterogeneous information systems,
biomedical ontologies were being developed \cite{Smith2007}. An
ontology is an explicit specification of a conceptualization of a
domain and can be used to make the meaning of terms in a vocabulary
explicit \cite{Gruber1995, Guarino1998}. They play a crucial role in
the annotation of biomedical data and the integration of model
organism databases \cite{go2010, Bada2004, goble}.

Ontologies increasingly rely on the use of Semantic Web technologies
\cite{Berners-Lee2001}. The Semantic Web provides a stack of protocols
and languages to include explicit semantics in websites. In
particular, the Web Ontology Language (OWL) \cite{Grau2008} has been
designed to express and share ontologies within the Semantic Web. OWL
is a language based on description logics (a group of formal languages
based on first-order predicate logic). Automated reasoners have been
developed within the Semantic Web to perform complex operations on
ontologies formulated in OWL. In particular, automated reasoners can
verify an ontology's consistency and use deductive inference to
perform powerful queries over ontologies. To benefit from automated
reasoning and the rapidly increasing number of software tools that are
being developed within the Semantic Web, most biomedical ontologies
are now available in OWL or can be converted into an OWL-based
representation \cite{Horrocks2007, Hoehndorf2010patterns}.

In the domain of phenotypes, multiple ontologies have been
developed. In particular, ontologies to characterize mammalian
\cite{Smith2004}, human \cite{Robinson2008}, yeast \cite{ypo} and worm
\cite{wpo} phenotypes are now available, with several more ontologies
being under development.

structures, qualities

open: process phenotypes very complex...  
physiology...
measurements and inferences...

@@

Figure \ref{fig:patterns} illustrates some examples of
non-comparative and comparative process attributes.

@@

While our considerations about process attributes are only the
beginnings of a full-fledged theory, we have derived several phenotype
formalization patterns and a high-level taxonomic structure of
process-based phenotypes. To evaluate our approach, we created the
Cellular Phenotype Ontology (CPO) by automatically applying our
patterns to the GO.

% CPO contains both structural abnormalities and physiological
% abnormalities of cells.
Phenotypes in the CPO are either based on structural abnormalities or
abnormal physiology involving cells or cell components. Structural
abnormalities in the CPO are based on GO's Cellular Component (GO-CC)
hierarchy. GO-CC contains 2,918 classes for cell parts (including {\em
  Cell}) and extracellular components of cells. For each cellular
component class $C$ in the GO-CC, we create a new class labelled {\em
  $C$ phenotype} in the CPO. For example, for the class {\em
  Mitochondrion} ({\tt GO:0005739}) in the GO-CC, we create the class
{\em Mitochondrion phenotype}.
